# Bayesian basics 1 {#sec-bb1}

These lecture notes cover topics from:

-   @mcelreath2020statistical Ch. 2, except 2.4.2-2.4.3
-   @kurz2021statistical [2.4.5](https://bookdown.org/content/4857/)

Note that @kurz2021statistical is a "translation" of @mcelreath2020statistical into tidyverse and brms code.

## Preliminaries {#sec-bb1-prelim}

Load libraries we will need:

```{r, cache=FALSE}
library(arm) # for the invlogit function
library(tidyverse)
library(brms)
```

::: callout-tip
## Practical note

If you have loaded `rethinking`, you need to detach it before using `brms`. See [Sec 4.3.1 of Kurz](https://bookdown.org/content/4857/geocentric-models.html).
:::

Load the `diatones` dataset from [*Regression Modeling for Linguistic Data*](https://osf.io/pnumg/):

```{r}
diatones <- read.csv("https://osf.io/tqjm8/download")
```

This dataset is described in Sec. 6.1.1 of @rmld-book,[^bayesian-basics-1-1] where you can learn more as needed. For the moment, all we need to know is:

[^bayesian-basics-1-1]: PDF available [here](https://github.com/msonderegger/rmld-v1.1/).

-   There are 130 observations: English noun/verb pairs (or "words") like "research", "combat".
-   All had noun and verb forms pronounced with final stress ("proTEST") in 1700.
-   `stress_shifted` indicates whether the noun form of the verb shifted stress between 1700 and 2005---e.g. "protest" and "debate" have `stress_shifted`=1, 0.

## Bayes' theorem

Bayes' theorem (or "rule") says that for any events $A$ and $B$:

$$
P(A|B) = \frac{P(B|A) P(A)}{P(B)}
$$

Here, $P(A|B)$ and $P(B|A)$ are conditional probabilities, and Bayes' theorem follows almost directly from the definition of conditional probabilities. We can think of $A$ and $B$ above as discrete random variables, and the same rule holds for continuous random variables (that is, their probability density functions).

-   **Note**: a method which uses Bayes rule is not (automatically) "Bayesian" (McElreath pp. 38-39 box).

It is customary to introduce Bayes' theorem with an example where $A$ and $B$ are discrete, like medical diagnostics, to develop intuition for how it works and can give unexpected results, even in a simple case. An exercise is below (to skip in class).

::: {#exr-bb1-1}
Open Chapter 5 of [*Seeing Theory*](https://seeing-theory.brown.edu/bayesian-inference/index.html#section3) and go to "Bayes Theorem". Consider a Covid-19 testing example, characteristic of the early days of the pandemic:

-   $P(D)$ = 0.05, $P(H)$ = 0.95

-   $P(+|D) = 0.99$, $P(-|H)=0.95$

That is, the test detects Covid near-perfectly in people who have the disease ("very high sensitivity"), and gives a few false positives for people who don't ("high specificity"). We assume that the *base rate*---the percentage of people who have Covid---is low (5%).

a.  What is the posterior probability $P(D|+)$---that is, the probability that someone has Covid if they test positive?
b.  Why might your answer seem surprising?
c.  Now suppose that the base rate is 0.25 (25% of the population is infected). What is $P(D|+)$ now?
:::

## Prior, likelihood, posterior

McElreath (Sec. 2.1-2.2) introduces key concepts of Bayesian inference using marble-counting and "globe tossing" examples. We will instead use a simple linguistic example, after first introducing notation:

-   We are interested in a *parameter*, $p$.[^bayesian-basics-1-2]
-   We observe *data* $D$ to estimate the parameters.
-   We have some knowledge about how likely different parameter values are, the *prior* (probability): $P(p)$
-   The *likelihood* is the probability of the data for a particular parameter value: $P(D|p)$.
-   The *posterior* (probability) is the plausibility of any parameter value ($p$) after seeing the data: $P(p|D)$.

[^bayesian-basics-1-2]: In more complex models, there will be many parameters.

Bayes' rule gives:

$$
P(p | D ) = \frac{P(D|p) P(p)}{P(D)}
$$

$P(D)$ is called the *average probability* or *marginal likelihood*, or other names.

Consider the `diatones` data:

-   $p_{\text{shift}}$ is the probability that a word of English shifts stress between 1700 and 2005.
-   The data is a sequence of 0 or 1 values.
    -   Technically: each is a measurement of a random variable $y$. The $i$th observation is $y_i$. In this data $y_1 = 1$, $y_2 = 0$, and so on.

```{r}
diatones %>% select(word, stress_shifted) %>% head()
```

```{r}
xtabs(~stress_shifted, data=diatones)
```

To build a Bayesian model we must specify the exact data, then fit the model to the data. It is useful to always first have a *data story* (McElreath 2.2), describing the *generative* process behind the data. This helps to write down the model.

In our case, the story is just a description of how the process leading to this sample:

0.  We have no prior knowledge about how likely English N/V pairs ("words") are to shift stress.
1.  The true proportion of words which shift stress is $p$.
2.  Every time we look up a new English word in the 1700 and 2005 dictionaries, we get a new 0 (stress didn't shift) or 1 (stress shifted) observation, $y$.
3.  Each word lookup is independent of the others.

(I have added a Step 0 for the prior, which McElreath doesn't include.)

We observe $N$ words, of which $k$ have shifted stress.

This lets us write down a probability model:

-   $p \sim \text{Uniform}(0,1)$
-   $k \sim \text{Binomial}(N, p)$

See [here](https://shiny.rit.albany.edu/stat/binomial/) what the binomial distribution looks like for different $N$ and $p$ values, if you're rusty.

This probability model means the pieces $P(p)$, $P(D|p)$, and $P(D)$ of the equation above are all defined: once we know the observed data ($D$), the posterior distribution automatically results (left-hand side of Bayes' rule above).

This is *Bayesian update*, which can be thought of either as:

-   An incremental process---update the posterior after $y_i$; this is the prior when we see $y_{i+1}$.

-   Seeing all $N$ observations at once---update the posterior once.

We can substitute the formula for binomial likelihood in to Bayes' rule in this case to give the mathematical formula for the posterior:

$$
P(p | k) = \frac{{n\choose{k}} p^{k}(1-p)^{n-k}}{P(k)}  \quad \text{for } 0 \le p \le 1
$$

(The prior here, $P(p)$, is *flat*, a.k.a. *uniform*: it is 1 when $0 \le p \le 1$ and 0 otherwise.)

It is common to just write this kind of Bayesian update without the denominator, which is only there to make sure the posterior distribution is a probability distribution (it sums to 1 over all possible values of $p$). This is written with the "proportional to" symbol:

$$
P(p | k) \propto {n\choose{k}} p^{k}(1-p)^{n-k}  \quad \text{for } 0 \le p \le 1
$$

If you find the analytic derivation or more details for this case helpful, see Sec [Sec. 2.2](https://vasishth.github.io/bayescogsci/book/ch-introBDA.html#sec-analytical) of @bdacs. (If you find the math confusing, don't worry.)

::: {#exr-bb1-2}
We will instead get a feel for Bayesian update visually, by playing with the *Seeing Theory* app [here](https://seeing-theory.brown.edu/bayesian-inference/index.html#section3).

This app shows a [*beta*](https://en.wikipedia.org/wiki/Beta_distribution) prior on $p$, which has parameters $\alpha$ and $\beta$. When $\alpha=\beta=1$, this is a uniform distribution (over $p \in [0,1]$).

Let $p=0.2$ for this example---meaning that in reality, 1/5 of words shifted stress.

With default settings, change $p=0.2$. Click "flip the coin" 5 times, and watch what happens. Do this again (resetting first) a few times.

a.  What is the same every time? (Does the posterior narrow? Shift right or left?)
b.  What is different every time?
:::

::: {#exr-bb1-3}
Now do the same exercise, with 50 flips (with $p=0.2$)---use "Flip 10 times"x5.
:::

::: {#exr-bb1-4}
Play around with $\alpha$ and $\beta$ to get priors which captures the intuition that:

a.  Values of $p$ further from 0 are less likely
b.  Values of $p$ further from 1 are less likely
c.  Values of $p$ which are below 0.05 or above 0.95 are very unlikely
:::

::: callout-tip
## Linguistics note

Each of these priors could be reasonable when studying different kinds of language variation and change.

For example, prior (b) could be applied for data from a *completed* sound change, to capture the intuition that most sound changes eventually apply to all possible words (the "Neogrammarian hypothesis"). Prior (c) could be applied to data from a sound change in progress.
:::

We will discuss soon how to actually choose priors. The goal for now is just to get comfortable thinking about the concept, and the effect they have on the final model (the posterior).

## Prior vs. posterior and sample size

Now let's try manually computing the posterior distribution, starting from a uniform prior for $p$, for our `diatones` example. (This is the case being simulated in the Seeing Theory app.)

The plot below shows the prior and the posterior, assuming we have just seen the first:

-   5 observations
-   10 observations
-   65 observations (50% of the data)
-   130 observations (all data)

We can do this without simulation, in this case, because the beta distribution is the [*conjugate prior*](https://towardsdatascience.com/conjugate-prior-explained-75957dc80bfb) for a binomial distribution—meaning that the posterior is also a beta distribution. You don't need to understand this code.

```{r}
# Beta is conjugate prior for binomial means:
# If you start with a Beta(\alpha, \beta) prior, 
# then observe $k$ $y=1$ observations of $N$ total, the posterior is:
# Beta(k+\alpha, N-k + \beta)

# For this case: 

#The numbers of observations with `stress_shifted`=1 in the first 5/10/65/130 observations are:
k_5 <- sum(diatones$stress_shifted[1:5])
k_10 <- sum(diatones$stress_shifted[1:10])
k_65 <- sum(diatones$stress_shifted[1:65])
k_130 <- sum(diatones$stress_shifted[1:130])

#set up a vector of p=0...1
p_vec <- seq(0,1, by=0.01)

# beta dist parameters for the prior
a <- 1
b <- 1

## ugly generation of posterior for each case -- should use a map function!
bind_rows(
  data.frame(p=p_vec, post=dbeta(p_vec, a,b), n=0),
  data.frame(p=p_vec, post=dbeta(p_vec, a+k_5,b+5-k_5), n=5),
  data.frame(p=p_vec, post=dbeta(p_vec, a+k_10,b+10-k_10), n=10),
  data.frame(p=p_vec, post=dbeta(p_vec, a+k_65,b+65-k_65), n=65),
    data.frame(p=p_vec, post=dbeta(p_vec, a+k_130,b+130-k_130), n=130)

) %>% mutate(n=factor(n)) %>% ## make the plot
  ggplot(aes(x=p, y=post)) + geom_line(aes(color=n)) +
  ylab("Posterior prob") 
```

Note how the posterior shifts a lot more in the first 10 observations than in the last 50% of observations.

<!-- ```{r} -->

<!-- #| echo: false -->

<!-- #|  -->

<!-- # If you start with a $Beta(\alpha, \beta)$ prior, then observe $k$ $y=1$ observations of $N$ total, the posterior is: -->

<!-- # Beta(k+\alpha, N-k + \beta) -->

<!-- # For this case: -->

<!-- #The numbers of observations with `stress_shifted`=1 in the first 5/10/65/130 observations are: -->

<!-- k_5 <- sum(diatones$stress_shifted[1:5]) -->

<!-- k_10 <- sum(diatones$stress_shifted[1:10]) -->

<!-- k_65 <- sum(diatones$stress_shifted[1:65]) -->

<!-- k_130 <- sum(diatones$stress_shifted[1:130]) -->

<!-- #set up a vector of p=0...1 -->

<!-- p_vec <- seq(0,1, by=0.01) -->

<!-- # beta dist parameters for the prior -->

<!-- a <- 3 -->

<!-- b <- 3 -->

<!-- ## ugly generation of posterior for each case -- should use a map function! -->

<!-- bind_rows( -->

<!--   data.frame(p=p_vec, post=dbeta(p_vec, a,b), n=0), -->

<!--   data.frame(p=p_vec, post=dbeta(p_vec, a+k_5,b+5-k_5), n=5), -->

<!--   data.frame(p=p_vec, post=dbeta(p_vec, a+k_10,b+10-k_10), n=10), -->

<!--   data.frame(p=p_vec, post=dbeta(p_vec, a+k_65,b+65-k_65), n=65), -->

<!--     data.frame(p=p_vec, post=dbeta(p_vec, a+k_130,b+130-k_130), n=130) -->

<!-- ) %>% mutate(n=factor(n)) %>% ## make the plot -->

<!--   ggplot(aes(x=p, y=post)) + geom_line(aes(color=n)) + -->

<!--   ylab("Posterior prob") -->

<!-- ``` -->

This example demonstrates how:

-   The posterior is a compromise between the prior and the likelihood
-   For smaller samples, the prior matters more
-   **For large enough samples, the likelihood dominates**

<!-- - Why is it always said that the marginal likelihood doesn't really matter? show plots without dividing by this.  THe *shape* is the same, just a matter of normalization. -->

Again, [Sec. 2.2 of BDA for CogSci](https://vasishth.github.io/bayescogsci/book/ch-introBDA.html#sec-analytical) shows a similar example in much more detail (and discusses what "conjugate prior" means).

In practice you rarely have to do Bayesian analyses "analytically", by manually calculating the posterior distribution, and it is impossible for more realistic models. It is worth seeing this one example to emphasize that at a low level, **every** Bayesian model is just doing this process: given a prior and a probability model, calculate the likelihood of the observed data, use it to compute the posterior.

This is different from frequentist methods: even for a linear regression, much less a mixed-effects model, it's often harder to conceptualize what fitting the model is "doing". The black magic part of Bayesian modeling is the MCMC methods used to actually fit the models (compute the posterior)---but this is conceptually separate.

::: {#exr-bb1-5}
a.  Redo the plot above, now using a different prior: the one you chose for "Values of $p$ which are below 0.05 or above 0.95 are very unlikely" above. This just requires changing `a` and `b` in the code.

b.  Re-generate the plot with these `a` and `b`. What do you notice, comparing the two plots for lower `n` (5, 10) and higher `n` (65, 130)?
:::

## Computing the posterior

For most realistic models it is not possible to manually compute the posterior, so an approximation method is used. McElreath makes extensive use in the first chapters of two methods:

-   *Grid approximation*
-   *Quadratic approximation*, closely related to the *MAP estimate* ("maximum a -posteriori")

We will not be using these, but they are worth knowing about---especially MAP, which often gives similar results to frequentist methods.

::: callout-tip
## Practical note

An important practical example is the [`blme` package](https://cran.r-project.org/web/packages/blme/index.html) for fitting `lme4`-style mixed-effects models. This package:

-   Redefines these models in a Bayesian framework
-   Uses weakly-informative priors (more on this later)
-   Returns MAP estimates of model parameters (fixed, random-effect coefficients).

This alone is enough to eliminate many model convergence woes with `lme4` models, and fitting a `blme` model is often faster than the equivalent `brms` model. It will be less accurate (as the MAP approximation is just that).
:::

### MCMC

Instead, we will follow [Kurz](https://bookdown.org/content/4857/) by immediately turning to a third method---Markov Chain Monte Carlo (MCMC)---which is more generally applicable.

For now we will just introduce MCMC at a high level, and will return for more discussion in a future week. (hopefully) return in more detail in a later week. There are various introductions online (post one you like!) if you'd like to go further, at varying levels of technical detail. A relevant McElreath blog post is [here](http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/).

The basic idea behind MCMC is that it is often much easier to simulate from a posterior distribution than to actually compute it. MCMC algorithms "walk around" the posterior distribution in a stochastic way, taking a sample of the parameters (here, $p$) at each step. "walk around" is defined such that **the sample approximates the posterior distribution**---once you wait long enough.

The simplest MCMC algorithm is called Metropolis-Hastings. MCMC is "easier to see than to understand" (McElreath). Open [this demo](https://chi-feng.github.io/mcmc-demo/), by Chi Feng, then:

-   Click "Random Walk Metropolis Hastings"
-   Set "target distribution" to "standard"

The underlying distribution here is a 2D Gaussian---a Gaussian hill. (If this is confusing, you can just focus on one dimension---its a Gaussian.) Note how the marginal distributions (on the $x$ and $y$ axes) get closer to normal distributions the longer you let the simulation run.

We will discuss some in class.

<!-- We will return to MCMC later, but since we will start using it immediately, there are some facts to be aware of (TODO: finish) -->

<!-- - we'll return to MCMC later, but since we're going to start using it immediately, some things to be aware of: -->

<!-- -- model output = samples of parameter values from the posterior. -->

<!-- -- to have a good model, our sample has to approximate the posterior well -- that means independence of where you started the chain, and enough iterations.  once sampling well, *independent* samples.   Without confidence in the sample, the model is garbage. -->

<!-- This is why: -->

<!-- - sampler has to "burn in" => usually discard the first N "warmup" samples.   -->

<!-- - "thin" = could discard every K samples, hope that this makes more independent -->

<!-- - always run multiple chains => *multiple cores* needed/v useful. -->

<!-- - run for at least several thousand. no magic number. -->

<!-- -- Various diagnostics to check the chains/model, which we'll get to soon.  ("Rhat", "ESS", etc. are related to these in brms output) -->

## brms: a first look

To estimate $p$ in a frequentist framework, we would fit a logistic regression with an intercept and no predictors:

```{r}
m1_glm <- glm(stress_shifted ~ 1, data=diatones, family=
                'binomial')
```

The estimated intercept, and its 95% confidence intervals are (in log-odds):

```{r}
coefficients(m1_glm)
confint(m1_glm)
```

These correspond in probability to:

```{r}
invlogit(coefficients(m1_glm))
invlogit(confint(m1_glm))
```

So the estimate is $\hat{p} =$ `r round(invlogit(coefficients(m1_glm)),2)` (95% CI: 0.11-0.23).

The equivalent model is fitted in `brms`, with a uniform prior on $p$, as follows:

```{r}
diatones_m1 <-
  brm(data = diatones,
      family = binomial(link = "identity"),
      stress_shifted | trials(1) ~ 0 + Intercept,
      prior(beta(1, 1), class = b, lb = 0, ub = 1)
  )
```

`brm()` is the workhorse function for fitting bayesian regression models. The `family` argument and the model formula together express the likelihood for $p$: this is a logistic regression on data where each $y$ is 0 or 1. The `prior` argument captures the uniform ($\text{Beta}(1,1)$) prior. (This particular model needs an unusual structure to make the prior work---when we do logistic regression we'll proceed differently.)

```{r}
summary(diatones_m1)
```

We will discuss the arguments, the messages during fitting, and the model output more soon. For now we focus on the posterior for $p$. A numeric summary is:

```{r}
posterior_summary(diatones_m1)
```

(Ignore the second row for now.) This shows the mean, SD, and 95% quantiles of the posterior distribution for $p$. Note how similar these are to the `glm`-derived values above.

The actual posterior can be visualized by extracting and plotting the model's sample from the posterior, using the `as_draws_df()` function from the posterior package.[^bayesian-basics-1-3]

[^bayesian-basics-1-3]: `as_draws_*` functions from this package extract posterior draws in various formats, like `as_draws_matrix()`, `as_draws_list()`. Here we need draws in dataframe format.

```{r}
as_draws_df(diatones_m1) %>%
  ggplot(aes(x = b_Intercept)) +
  geom_density(fill = "black") +
  scale_x_continuous("Prob. of stress shifting", limits = c(0, 1)) + 
  ylab("Posterior probability")
```

(Usually we will just call this "plotting the results", or similar.) This looks very similar to the analytically-computed posterior above (for all $n=130$ observations).

::: {#exr-bb1-6}
a.  Refit the `brms` model, now with your prior that captures that values of $p$ very near 0 or 1 are unlikely. How does the posterior look different, if at all?

<!-- (A: it shouldn't) -->

b.  Now refit both of the last two models, but just using the first 20 observations from `diatones`. How do the posteriors of these models differ? Why? <!-- (A: it should. fewer samples) -->
:::

## Solutions

### @exr-bb1-4

a.  $\alpha = 1, \beta = 3$

b.  $\alpha = 3, \beta = 1$

c.  $\alpha = 2, \beta = 2$

The exact $\alpha/\beta$ values here matter less than the qualitative pattern: $\alpha < \beta$, $\alpha > \beta$, and $\alpha = \beta$ with $\alpha = \beta > 1$.

### @exr-bb1-6

a.  Refit the brms model, now with your prior that captures that values of very near 0 or 1 are unlikely. How does the posterior look different, if at all?

```{r}
diatones_m1_newprior <-
  brm(data = diatones,
      family = binomial(link = "identity"),
      stress_shifted | trials(1) ~ 0 + Intercept,
      prior(beta(2, 2), class = b, lb = 0, ub = 1),
        file = "models/diatones_m1_newprior.brm"
  )

diatones_m1_newprior
```

Visualize posterior:

```{r}
as_draws_df(diatones_m1_newprior) %>%
  ggplot(aes(x = b_Intercept)) +
  geom_density(fill = "black") +
  scale_x_continuous("Prob. of stress shifting", limits = c(0, 1)) + 
  ylab("Posterior probability")

```

This looks very similar to the `diatones_m1` posterior, despite a different prior: the likelihood has overwhelmed the prior.

b.  

```{r}

diatones_m1_20 <-
  brm(data = diatones[1:20,],
      family = binomial(link = "identity"),
      stress_shifted | trials(1) ~ 0 + Intercept,
      prior(beta(1, 1), class = b, lb = 0, ub = 1),
      file = "models/diatones_m1_20.brm"
  )


diatones_m1_newprior_20 <-
  brm(data = diatones[1:20,],
      family = binomial(link = "identity"),
      stress_shifted | trials(1) ~ 0 + Intercept,
      prior(beta(2, 2), class = b, lb = 0, ub = 1),
        file = "models/diatones_m1_newprior_20.brm"
  )
```

```{r, fig.width=4, fig.height=3}
#| layout-ncol: 2
as_draws_df(diatones_m1_20) %>%
  ggplot(aes(x = b_Intercept)) +
  geom_density(fill = "black") +
  scale_x_continuous("Prob. of stress shifting", limits = c(0, 1)) + 
  ylab("Posterior probability") + ggtitle("Flat prior")


as_draws_df(diatones_m1_newprior_20) %>%
  ggplot(aes(x = b_Intercept)) +
  geom_density(fill = "black") +
  scale_x_continuous("Prob. of stress shifting", limits = c(0, 1)) + 
  ylab("Posterior probability") + ggtitle("Non-flat prior")

```

The posterior is slightly narrower for the second model, with some probability weight shifted "to the right" away from 0, as we expect given the prior. The effect of the prior is now (barely) visible because there is less data, so the prior is not quite overwhelmed.
